---
title: "Which extreme weather events are most harmful"
output: html_document
---

## Synopsis
Extreme weather events may be very harmful both for the public health and economics.
Tornadoes are consistently one of the most harmful events, every year hundreds of people across US are killed and thousands injured.
Until 1992 only tornados were recorded. 
After 1992 other extreme weather events were reported, and turned out to be even more dangerous.
Excessive heat caused several times more fatalities that tornados, floods cause more injuries,
and storm winds cause more property damage

## Data Processing

Read the raw data
``` {r cache=TRUE}
rawDataFile <- "stormData.zip"
if (!file.exists(rawDataFile)) {
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", rawDataFile);
}
rawData <- read.csv(rawDataFile);
```

Build a new data set that includes only the events that caused fatalities, injuries, or damage
``` {r}

harm <- rawData[rawData$FATALITIES > 0 | 
                    rawData$INJURIES > 0 | 
                    rawData$PROPDMG > 0, 
                c("EVTYPE", "BGN_DATE", "FATALITIES", "INJURIES", "PROPDMG")]
harm$BGN_DATE <- as.Date(harm$BGN_DATE, format = "%m/%d/%Y")
harm$YEAR <- as.numeric(format(harm$BGN_DATE, "%Y"))
```

Pick the most significant event types according to the following logic:  
* calculate sum of particular type of damage (fatalities, injuries, property damage)
for each event type across all the years.  
* for each event type calculate its percentage in the total damage
* choose the event types with the highest percentage  
* only TORNADO events were reported until 1992, but this only affects the percentage of
TORNADO, it is one of the most significant factors even after 1992.

The following function implements the calculation of *num* most significant event types
for the given *field* across all the years
``` {r}
topEventTypes <- function(field, num) {
  rows <- harm[, field] > 0
  res <- aggregate(list(COUNT = harm[rows, field]), list(EVTYPE = harm[rows, "EVTYPE"]), sum)
  total <- sum(harm[rows, field])
  res$PERCENT <- round(res$COUNT/total * 100)  
  res[order(-res$COUNT),][1:num,]
}
```

Most harmful events for FATALITIES
``` {r}
fatalityEvents <- topEventTypes("FATALITIES", 10)
fatalityEvents[,c("EVTYPE", "PERCENT")]
```

Most harmful events for INJURIES
``` {r}
injuryEvents <- topEventTypes("INJURIES", 10)
injuryEvents[,c("EVTYPE", "PERCENT")]
```

Most harmful events for PROPDMG
``` {r}
propEvents <- topEventTypes("PROPDMG", 10)
propEvents[,c("EVTYPE", "PERCENT")]
```

Concatenate the top 5 event types from all damage types, 
and assign a color to each event type so that they could be plotted consistently
``` {r}
eventTypes <- unique(c(fatalityEvents$EVTYPE[1:5], injuryEvents$EVTYPE[1:5], propEvents$EVTYPE[1:5]))
eventTypes <- factor(eventTypes, levels=1:nlevels(harm$EVTYPE), labels=levels(harm$EVTYPE))
eventColors <- data.frame(EVTYPE = eventTypes, COL = c("red", "blue", "green", "orange", "yellow", "brown", "cyan", "magenta"))
eventColors$COL <- as.character(eventColors$COL)
eventColors
```

Now let's produce a plot for each damage type.
``` {r}
plotDamage <- function(field, evTypes, title) {
    colors <- c()
    for (t in evTypes) {
        colors <- c(colors, eventColors[eventColors$EVTYPE == t, "COL"])
    }

    events <- aggregate(list(COUNT = harm[,field]), list(YEAR = harm$YEAR, EVTYPE = harm$EVTYPE), sum)

    plot(events$YEAR, events$COUNT, type = "n", main = title, xlab = "Year", ylab = field)
    legend("topleft", legend = evTypes, lty = c(1,1), col = colors)

    for (type in evTypes) {
        color <- eventColors[eventColors$EVTYPE == type, "COL"]
        lines(events[events$EVTYPE == type, "YEAR"], events[events$EVTYPE == type, "COUNT"], col=color)
    }
}
```

``` {r}
plotDamage("FATALITIES", fatalityEvents$EVTYPE[1:5], "Fatalities by event types")
```

``` {r}
plotDamage("INJURIES", injuryEvents$EVTYPE[1:5], "Injuries by event types")
```

``` {r}
plotDamage("PROPDMG", propEvents$EVTYPE[1:5], "Property damages by event types")
```

## Results

Until 1995 tornadoes caused most damages, both to the public health and economics.
From 1995 heat and flood caused much more damages than tornadoes. 

### Fatalities
TORNADO was the main and almost only reason for the fatalities until 1995.
However in the period from 1995 to 2010 HEAT and EXCESSIVE HEAT have caused much more fatalities than.
In 2011 TORNADO again becomes the most fatal event

### Injuries
TORNADO consistently causes the most injuries from the beginning of recording.
Only in period from 1995 to 2007 FLOOD causes the most injuries, and EXCESSIVE HEAT becomes as dangerous as TORNADO

### Property damage
TORNADO caused all the damages in the period from 1950 to 1995.
From 1995 to 2010 HEAT and EXCESSIVE HEAT cause much more damages, and FLASH FLOOD caused as much damage as TORNADO